---
title: "STAT6 / CRSwNP Receptor & Signaling Analysis"
subtitle: "Type 1/2 Receptor Expression, Ligand Expression, Cell-Cell Communication,
  and Downstream STAT6 Readouts"
date: "2026-07"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  root.dir = "~/project/results_nofilter_gene/integration_harmony/",
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, fig.align = "center", fig.pos = "H"
)

fig_dir <- "~/project/results_nofilter_gene/integration_harmony/STAT6/"

.libPaths("~/R/intern")
options(future.globals.maxSize = 2 * 1024^3)  # 2 GB
library(Seurat)
library(ggplot2)
library(dplyr)
library(harmony)
library(patchwork)
library(tidyverse)
library(Nebulosa)
theme_set(theme_bw(base_size = 11))
```

# Overview

This document implements the analysis plan addressing four questions:

1. Which cell types express Type 1 (IL4R + IL2RG) vs Type 2 (IL4R + IL13RA1) receptors, and are there differences across conditions (Control / CRSsNP_Eth / CRSwNP_Eth / CRSwNP_NP)?
2. Which cell types express the ligands IL4 and IL13, and are there differences across conditions?
3. What can we say about cell-cell communication for IL4/IL13 signaling in CRSwNP vs Control/CRSsNP?
4. Which cell types express downstream pSTAT6 readouts (CCL17, CCL26, etc.), and is there a global increase in CRSwNP?

# Data Input

```{r}
obj <- readRDS('~/project/results_nofilter_gene/integration_harmony/GSE235711_CRS_annotated_seurat_obj.rds')
c('IL4R', 'IL2RG', 'IL13RA1') %in% row.names(obj)
```

\newpage

# Type 1 vs Type 2 Receptor Expression

```{r}
expr <- GetAssayData(
  obj,
  assay = "RNA",
  slot = "data"
)[c("IL4R", "IL2RG", "IL13RA1"), ]

# Create dataframe
df <- data.frame(
  condition = obj$condition2,
  sample_id = obj$sample_id,
  #celltype = obj$celltype3,
  celltype = obj$celltype,
  IL4R = as.numeric(expr["IL4R", ]),
  IL2RG = as.numeric(expr["IL2RG", ]),
  IL13RA1 = as.numeric(expr["IL13RA1", ])
)

# Average expression per sample and cell type
df <- df %>%
  group_by(condition, celltype, sample_id) %>%
  summarise(
    IL4R = mean(IL4R),
    IL2RG = mean(IL2RG),
    IL13RA1 = mean(IL13RA1),
    n_cells = n(),
    .groups = "drop"
  )

summary_df <- df %>%
  group_by(celltype) %>%
  summarise(min_IL4R = min(IL4R), median_IL4R = median(IL4R),
            min_IL2RG = min(IL2RG), median_IL2RG = median(IL2RG),
            min_IL13RA1 = min(IL13RA1), median_IL13RA1 = median(IL13RA1))

df <- df %>% left_join(summary_df, by = "celltype")
```

## Co-expression by cell type

```{r}
ggplot(df, aes(x = IL4R, y = IL2RG)) +
  geom_point(
    aes(
      size = (log10(n_cells)),
      color = condition
    ),
    alpha = 0.8
  ) +
  facet_wrap(~celltype, scales = "free_y") +
  scale_size_area(max_size = 5, name = "log10(Number of cells)") +
  labs(
    x = "Mean Log-normalized IL4R expression",
    y = "Mean Log-normalized mean IL2RG expression",
    color = "Condition"
  ) +
  theme_bw()+
  coord_cartesian(ylim = c(0, 1))
```

```{r}
ggplot(df, aes(x = IL4R, y = IL13RA1)) +
  geom_point(
    aes(
      size = log10(n_cells),
      color = condition
    ),
    alpha = 0.8
  ) +
  facet_wrap(~celltype, scales = "free_y") +
  scale_size_area(max_size = 5, name = "log10(Number of cells)")+
  labs(
    x = "Mean Log-normalized IL4R expression",
    y = "Mean Log-normalized mean IL13RA1 expression",
    color = "Condition"
  ) +
  coord_cartesian(ylim = c(0, 1))+
  theme_bw()
```


```{r}
my_cols <- c(
  "#fb6f92", "#c095e4", "#008585", "#ffa600", "#43b0f1",
  "#a96b6c", "#ffca3a", "#92ba92", "#ff6361", "#eac4d5"
)


global_IL4R_mean   <- mean(expr["IL4R", ])
global_IL2RG_mean <- mean(expr["IL2RG", ])
global_IL13RA1_mean <- mean(expr["IL13RA1", ])


p1 <- ggplot(
  bubble_df,
  aes(
    IL4R,
    IL2RG,
    size = log10(n_cells),
    color = celltype
  )
) +
  geom_vline(
    xintercept = global_IL4R_mean,
    linetype = "dashed",
    color = "blue"
  ) +
  geom_hline(
    yintercept = global_IL2RG_mean,
    linetype = "dashed",
    color = "blue"
  ) +
  geom_point(alpha = 0.9) +
  geom_text(
    aes(label = celltype),
    color = "black",
    size = 3.5,
    vjust = -0.5,
    check_overlap = TRUE,
    show.legend = FALSE
  ) +
  scale_size_continuous(
    range = c(3, 8),
    name = "log10(Cell count)"
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0.15, 0.05))
  ) +
  scale_color_manual(values = my_cols) +
  guides(color = "none") +
  theme_classic() +
  labs(
    x = "Mean IL4R expression",
    y = "Mean IL2RG expression"
  )

p2 <- ggplot(
  bubble_df,
  aes(
    IL4R,
    IL13RA1,
    size = log10(n_cells),
    color = celltype
  )
) +
  geom_vline(
    xintercept = global_IL4R_mean,
    linetype = "dashed",
    color = "blue"
  ) +
  geom_hline(
    yintercept = global_IL13RA1_mean,
    linetype = "dashed",
    color = "blue"
  ) +
  geom_point(alpha = 0.9) +
  geom_text(
    aes(label = celltype),
    color = "black",
    size = 3.5,
    vjust = -0.5,
    check_overlap = TRUE,
    show.legend = FALSE
  ) +
  scale_size_continuous(
    range = c(3, 8),
    name = "log10(Cell count)"
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0.15, 0.05))
  ) +
  scale_color_manual(values = my_cols) +
  guides(color = "none") +
  theme_classic() +
  labs(
    x = "Mean IL4R expression",
    y = "Mean IL13RA1 expression"
  )

print(p1)
print(p2)
```


## "IL4R", "IL2RG", "IL13RA1" Dotplot, FeaturePlot & Density Plot
```{r}
DotPlot(
  object = obj,
  features = c("STAT6", "IL4R", "IL2RG", "IL13RA1"),
  group.by = "celltype3",
  scale = F)

FeaturePlot(obj, 
            reduction = "umap.harmony",
            features = c("STAT6", "IL4R", "IL2RG", "IL13RA1")
)
```

```{r}
p <- Nebulosa::plot_density(
    obj,
    features = "STAT6",
    reduction = "umap.harmony"
)
p$labels$colour <- "Density"
p +
  theme_classic() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "grey60"),
    axis.text = element_text(color = "black"),
    axis.title = element_text(color = "black")
  )

```

```{r}
p <- Nebulosa::plot_density(
    obj,
    reduction = "umap.harmony",
    "IL4R"
)
p$labels$colour <- "Density"
p
```

```{r}
p <- Nebulosa::plot_density(
    obj,
    reduction = "umap.harmony",
    "IL2RG"
)
p$labels$colour <- "Density"
p
```

```{r}
p <- Nebulosa::plot_density(
    obj,
    reduction = "umap.harmony",
    "IL13RA1"
)
p$labels$colour <- "Density"
p
```

## Double-positive proportions by cell type
```{r}
expr <- FetchData(obj, vars = c('IL4R', 'IL2RG', 'IL13RA1', 'celltype3', 'condition2', 'sample_id'), slot = 'data')
expr %>%
  group_by(celltype3) %>%
  summarise(
    pct_type1_pos = mean(IL4R > 0 & IL2RG > 0) * 100,
    pct_type2_pos = mean(IL4R > 0 & IL13RA1 > 0) * 100,
    n_cells = n()
  )
```

## Type 1 receptor double-positive heatmap

```{r}
receptor_summary <- expr %>%
  group_by(celltype3, condition2) %>%
  summarise(
    pct_type1_pos = mean(IL4R > 0 & IL2RG > 0) * 100,
    pct_type2_pos = mean(IL4R > 0 & IL13RA1 > 0) * 100,
    n_cells = n()
  )
```

```{r}
ggplot(receptor_summary, aes(x = condition2, y = celltype3, fill = pct_type1_pos)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(pct_type1_pos, 1)), size = 3) +
  scale_fill_gradient(low = "white", high = "darkred", name = "% double+\n(Type 1)") +
  labs(title = "Type 1 Receptor (IL4R+IL2RG) Double-Positive %", x = "", y = "")
```

```{r}
ggplot(receptor_summary, aes(x = condition2, y = celltype3, fill = pct_type2_pos)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(pct_type2_pos, 1)), size = 3) +
  scale_fill_gradient(low = "white", high = "darkred", name = "% double+\n(Type 2)") +
  labs(title = "Type 2 Receptor (IL4R+IL13RA1) Double-Positive %", x = "", y = "")
```

## Sample-level comparison across conditions

```{r}
sample_level <- expr %>%
  group_by(celltype3, condition2, sample_id) %>%
  summarise(
    pct_type1_pos = mean(IL4R > 0 & IL2RG > 0) * 100,
    pct_type2_pos = mean(IL4R > 0 & IL13RA1 > 0) * 100,
    n_cells = n(),
    .groups = "drop"
  ) %>%
  filter(n_cells >= 20)
```

```{r}
ggplot(sample_level, aes(x = condition2, y = pct_type1_pos, color = condition2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.15, size = 1.8) +
  #facet_wrap(~celltype3, scales = "free_y", ncol = 4) +
  facet_wrap(~celltype3, ncol = 4) +
  labs(y = "% IL4R+IL2RG double-positive cells", x = "") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        strip.text = element_text(size = 7))
```

```{r}
ggplot(sample_level, aes(x = condition2, y = pct_type2_pos, color = condition2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.15, size = 1.8) +
  facet_wrap(~celltype3, ncol = 4) +
  labs(y = "% IL4R+IL13RA1 double-positive cells", x = "") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        strip.text = element_text(size = 7))
```

\newpage

# Ligand (IL4/IL13) Expression

```{r}
expr <- FetchData(obj, vars = c('IL4', 'IL13', 'celltype3', 'condition2', 'sample_id'), slot = 'data')
ligand_summary <- expr %>%
  group_by(celltype3, condition2) %>%
  summarise(
    pct_IL4_pos  = mean(IL4  > 0) * 100,
    pct_IL13_pos = mean(IL13 > 0) * 100,
    mean_IL4  = mean(IL4),
    mean_IL13 = mean(IL13),
    .groups = "drop"
  )
ligand_summary
```

## DotPlot-style ligand expression by cell type and condition

```{r}
dotplot_df <- ligand_summary %>%
  select(celltype3, condition2, pct_IL4_pos, pct_IL13_pos, mean_IL4, mean_IL13) %>%
  pivot_longer(cols = c(pct_IL4_pos, pct_IL13_pos, mean_IL4, mean_IL13),
                names_to = "key", values_to = "value") %>%
  mutate(gene = ifelse(grepl("IL4$|IL4_", key), "IL4", "IL13"),
          metric = ifelse(grepl("^pct", key), "pct_pos", "mean_expr")) %>%
  select(-key) %>%
  pivot_wider(names_from = metric, values_from = value)

ggplot(dotplot_df, aes(x = gene, y = celltype3, size = pct_pos, color = mean_expr)) +
  geom_point() +
  facet_wrap(~condition2, nrow = 1) +
  scale_size_continuous(range = c(0, 8), name = "% expressing") +
  scale_color_gradient(low = "lightgrey", high = "darkblue", name = "Mean expr.") +
  labs(title = "IL4/IL13 Ligand Expression Across Cell Types and Conditions", x = "", y = "") +
  theme(axis.text.x = element_text(angle = 0))

# Desired order (top to bottom)
celltype_order <- c(
  "MCT",
  "MCT_CTSG",
  "MCTC",
  "Mono_Macro",
  "Monocytes",
  "Macrophages",
  "pDCs",
  "Neutrophil",
  "T/NK cells",
  "B cells",
  "Plasma cells",
  "Fibroblasts",
  "Endothelial cells",
  "Epithelial cells"
)

dotplot_df$celltype <- factor(
  dotplot_df$celltype3,
  levels = rev(celltype_order)  # reverse because ggplot draws first level at bottom
)
ggplot(dotplot_df, aes(x = condition2, y = celltype, size = pct_pos, color = mean_expr)) +
  geom_point() +
  facet_wrap(~gene, nrow = 1) +
  scale_size_continuous(range = c(0, 8), name = "% expressing") +
  scale_color_gradient(low = "lightgrey", high = "darkblue", name = "Mean expr.") +
  labs(title = "IL4/IL13 Ligand Expression Across Cell Types and Conditions", x = "", y = "") +
  theme(axis.text.x = element_text(
    angle = 45,
    hjust = 1,
    vjust = 1
))
```

\newpage

# Downstream STAT6 Readouts (CCL17, CCL26, AMICA1, GATA3)

## Pseudobulk CCL17 expression by condition

```{r}
c('CCL17', 'CCL26', 'AMICA1', 'GATA3', 'STAT6') %in% rownames(obj)
# 'AMICA1' not find in this dataset
expr <- FetchData(obj, vars = c('CCL17', 'CCL26', 'GATA3', 'STAT6', 'celltype3', 'condition2', 'sample_id'), slot = 'data')
pseudobulk <- expr %>%
  group_by(sample_id, condition2) %>%
  summarise(STAT6 = mean(STAT6),
            CCL17 = mean(CCL17), 
            CCL26 = mean(CCL26), 
            GATA3 = mean(GATA3), 
            .groups = "drop")
pseudobulk_long <- pseudobulk %>%
  pivot_longer(
    cols = c(STAT6, CCL17, CCL26, GATA3),
    names_to = "gene",
    values_to = "expression"
  )
pseudobulk_long$gene <- factor(
  pseudobulk_long$gene,
  levels = c("STAT6", "CCL17", "CCL26", "GATA3")
)

ggplot(pseudobulk_long,
       aes(x = condition2, y = expression, color = condition2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.15, size = 3) +
  #facet_wrap(~gene, scales = "free_y", nrow = 1) +
  facet_wrap(~gene, nrow = 1) +
  labs(
    title = "",
    x = "",
    y = "Mean log-normalized expression"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## STAT6/CCL17/CCL26/GATA3 by cell type and condition (DotPlot-style)

```{r}
target_summary <- expr %>%
  group_by(celltype3, condition2) %>%
  summarise(
    pct_STAT6 = mean(STAT6 > 0) * 100, mean_STAT6 = mean(STAT6),
    pct_CCL17 = mean(CCL17 > 0) * 100, mean_CCL17 = mean(CCL17),
    pct_CCL26 = mean(CCL26 > 0) * 100, mean_CCL26 = mean(CCL26),
    pct_GATA3 = mean(GATA3 > 0) * 100, mean_GATA3 = mean(GATA3),
    .groups = "drop"
  )

target_summary$celltype3 <- factor(
  target_summary$celltype3,
  levels = rev(celltype_order)  # reverse because ggplot draws first level at bottom
)

dotplot_df2 <- target_summary %>%
  pivot_longer(
    -c(celltype3, condition2),
    names_to = "key",
    values_to = "value"
  ) %>%
  mutate(
    gene = sub("^(pct|mean)_", "", key),
    metric = ifelse(grepl("^pct_", key), "pct_pos", "mean_expr")
  ) %>%
  select(-key) %>%
  pivot_wider(
    names_from = metric,
    values_from = value
  )

dotplot_df2$gene <- factor(
  dotplot_df2$gene,
  levels = c("STAT6", "CCL17", "CCL26", "GATA3")
)

ggplot(
  dotplot_df2,
  aes(
    x = condition2,
    y = celltype3,
    size = pct_pos,
    color = mean_expr
  )
) +
  geom_point() +
  facet_wrap(~gene, nrow = 1) +
  scale_size_continuous(
    range = c(0, 8),
    name = "% expressing"
  ) +
  scale_color_gradient(
    low = "lightgrey",
    high = "darkblue",
    name = "Mean expr."
  ) +
  labs(
    title = "STAT6 and STAT6-Related Downstream Genes Expression",
    x = "",
    y = ""
  )+
  theme_bw() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```